---
title: "1-permafrost-data-prep"
author: "bbl"
format:
  html:
    embed-resources: true
    code-fold: true
toc: true
editor: visual
---

**TODO:**

- Check that pre-199x data aren't getting PF data
- Try global Random Forest model with permafrost flag
- CLEAN UP GRACE'S STUFF AND MERGE AND UPDATE
- Deal with the remaining papers in my to-process folder, prioritizing high latitude ones
- Find and fix NA biomes
- Read about VIF

This script does prep work for the main analysis asking whether
permafrost *presence* and *thermal state* affect the measured
soil-to-atmosphere CO2 flux ("soil respiration").

```{r setup}
#| include: false

# Data processing and visualization packages
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())
library(lubridate)
library(parallel)

# File IO packages
library(ncdf4)
library(arrow)
library(readr)

# Statistics and spatial packages
library(broom)
library(sf)
library(terra)

# Data packages
library(geodata)
library(rnaturalearth)
library(MODISTools)
```

# SRDB

The Soil Respiration Database (SRBD) is described by 
[Jian et al. (2021)](https://essd.copernicus.org/articles/13/255/2021/),
but the version we use here has had extensive updates since then.

```{r read-srdb}
#| fig-width: 12

read_csv("data-srdb/srdb-data.csv", show_col_types = FALSE) |>  
    select(Record_number, Study_number, Author, 
           Year = Study_midyear, YearsOfData, 
           Latitude, Longitude, Elevation, Manipulation, 
           Biome, Ecosystem_state, 
           Rs_annual, Rh_annual, Rs_growingseason,
           Ecosystem_type,
           # intercomparison columns -- will not be used afterwards
           SRDB_study_temp = Study_temp, 
           SRDB_study_precip = Study_precip, 
           SRDB_GPP = GPP, 
           SRDB_leaf_habit = Leaf_habit) |>  
    filter(Manipulation == "None") |> 
    filter(!Ecosystem_type %in% c("Agriculture", "Bare", "Urban")) |> 
    filter(!is.na(Longitude), !is.na(Latitude), !is.na(Year)) |> 
    filter(!is.na(Rs_growingseason) | 
             !is.na(Rs_annual) | 
             !is.na(Rh_annual)) |> 
    select(-Elevation, -Manipulation) |> 
    mutate(Longitude = round(Longitude, 4),
           Latitude = round(Latitude, 4),
           Year = floor(Year)) ->
    srdb

message(nrow(srdb), " rows of data")

world <- ne_coastline()
srdb_annual <- srdb |> filter(is.na(Rs_annual))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_annual, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rs_annual N=", nrow(srdb_annual)))

srdb_rh_annual <- srdb |> filter(!is.na(Rh_annual))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_rh_annual, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rh_annual N=", nrow(srdb_rh_annual)))

srdb_growingseason <- srdb |> filter(is.na(Rs_growingseason))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_growingseason, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rs_growingseason N=", nrow(srdb_growingseason)))
```

```{r srdb-prep}

# Subsample SRDB data for quick testing
set.seed(123)
# srdb |>
#   filter(Latitude > 50) |> 
#     slice_sample(n = 100) ->
#     srdb_sample
srdb_sample <- srdb

srdb_sample |> 
    select(Record_number, Longitude, Latitude) ->
    coords

srdb_sample |> 
  distinct(Year, Longitude, Latitude) ->
  coords_yr
```

# ESA-CCI permafrost

Our data source for permafrost temperatures is Westermann et al.: 
ESA permafrost Climate Change Initiative (permafrost_cci): 
Permafrost ground temperature for the Northern Hemisphere, v5.0,
<https://doi.org/10.5285/5675B0BE944F45A8AF0E7DDBEB47A011>, 2025.

```{r read-essacci-permafrost}
#| fig-height: 6
#| fig-width: 10

PERMAFROST_CACHE <- "caches/permafrost_cache.parquet"
if(file.exists(PERMAFROST_CACHE)) {
  message("Reading data from permafrost cache...")
  pf_cache <- arrow::read_parquet(PERMAFROST_CACHE) 
  message("\t(", nrow(pf_cache), " rows)")
} else {
  message("No permafrost cache file found")
  pf_cache <- tibble(Record_number = numeric(0), 
                     Longitude = numeric(0), 
                     Latitude = numeric(0))
}

# Isolate SRDB lon/lats NOT in the cache
coords |> 
  anti_join(pf_cache, by = c("Record_number", "Longitude", "Latitude")) ->
  coords_no_data
if(nrow(coords_no_data) > 0) {
  message("There are ", nrow(coords_no_data), 
          " coordinates that need data")
  
  results_list <- list()

  files <- list.files("data-permafrost_cci//", 
                      full.names = TRUE,
                      pattern = "\\.nc$",
                      recursive = TRUE)
 
  for(fn in files) {
    # Trying a simple terra::rast() with these data gives
    # Error: [rast] could not load multidimensional data. 
    # So use ncdf4 to load; transpose and flip; and convert
    # to a raster, specifying the spatial extent and CRS
    message("Opening ", basename(fn))
    nc <- nc_open(fn)
    
    ncvars_to_read <- c("GST", "GST_uncertainty", "T1m", "T1m_uncertainty")
    for(v in ncvars_to_read) {
      message("\tGetting ", v)
      dat_matrix <- ncvar_get(nc, varid = v)
      
      # Transpose and flip the data
      dat_matrix <- t(dat_matrix)
      dat_matrix <- dat_matrix[dim(dat_matrix)[1]:1,]
      # Convert to spatial raster, specifying lon/lat extent
      dat_rast <- rast(dat_matrix, 
                       extent = c(-180, 180, 25, 85), 
                       crs = "+proj=longlat +datum=WGS84")
      message("\tExtracting points...")
      xdat <- terra::extract(dat_rast,
                             coords_no_data[c("Longitude", "Latitude")])
      coords_no_data[paste0("PF_", v)] <- xdat[2]
    }
    rm(dat_matrix, dat_rast) # clean up; these are big
    
    # The year is in the filename; extract it (a bit hackily)
    yr <- strsplit(basename(fn), "-", fixed = TRUE)[[1]][7]
    coords_no_data$Year <- as.numeric(yr)
    results_list[[yr]] <- coords_no_data
    
    nc_close(nc)
  }
  
  bind_rows(results_list) |> 
    bind_rows(pf_cache) ->
    pf_cache
  
  message("Writing cache file (", nrow(pf_cache), " rows)...")
  arrow::write_parquet(pf_cache, PERMAFROST_CACHE)
}

pf_cache |> 
  filter(!is.na(PF_T1m)) ->
  pf_cache_good

ggplot(pf_cache_good,
       aes(Year, PF_T1m, color = Latitude, group = Record_number)) +
  geom_line() + 
  ggtitle("1m permafrost temperatures over time")

ggplot(pf_cache_good,
       aes(PF_GST, PF_T1m, color = Latitude)) +
  geom_point(size = 0.5) + 
  ggtitle("Surface versus 1m permafrost temperatures")

pf_cache_good |> 
  distinct(Record_number, Year, Longitude, Latitude) ->
  pf_points
```

Dataset | N 
 -- | --
SRDB entries | `r nrow(srdb_sample)`
SRDB geographic points | `r nrow(coords_yr |> distinct(Longitude, Latitude))`
SRDB entries with permafrost| `r nrow(pf_cache_good)`
SRDB year-geographic points with permafrost| `r nrow(pf_cache_good |> distinct(Latitude, Longitude))`

## Permafrost trends

For each geographic point, compute mean, trend, and p-value of that trend.

```{r permafrost-stats}
#| fig-width: 12

message("Computing means and trends...")
pf_cache_good |> 
  select(Record_number, Longitude, Latitude, Year, starts_with("PF_")) |> 
  select(-ends_with("uncertainty")) ->
  results_non_uncertainty

# Compute means
results_non_uncertainty |> 
  pivot_longer(starts_with("PF_")) |> 
  group_by(Record_number, Longitude, Latitude, name) |> 
  summarise(value = mean(value), .groups = "drop") |> 
  mutate(name = paste0(name, "_mean")) |> 
  pivot_wider() ->
  results_mean

# Compute trends
results_non_uncertainty |> 
  pivot_longer(starts_with("PF_")) |> 
  filter(!is.na(value)) |> 
  group_by(Record_number, Longitude, Latitude, name) |> 
  reframe(lm(value ~ Year) |> tidy()) |> 
  filter(term == "Year") |> 
  mutate(estimate = estimate * 10) |> # change to degrees per decade
  select(Record_number, Longitude, Latitude, name, estimate, p.value) ->
  trends

trends |> 
  select(Record_number, Longitude, Latitude, name, estimate) |> 
  mutate(name = paste0(name, "_trend")) |> 
  pivot_wider(values_from = "estimate") ->
  trends_est
trends |> 
  select(Record_number, Longitude, Latitude, name, p.value) |> 
  mutate(name = paste0(name, "_p.value")) |> 
  pivot_wider(values_from = "p.value") ->
  trends_p

results_mean |> 
  left_join(trends_est, 
            by = c("Record_number", "Longitude", "Latitude")) |> 
  left_join(trends_p, 
            by = c("Record_number", "Longitude", "Latitude")) ->
  pf_stats

# Visualize

ggplot(world) + 
    geom_sf() + 
    geom_point(data = pf_stats, 
               aes(x = Longitude, y = Latitude, 
                   color = PF_T1m_mean)) + 
  coord_sf(ylim = c(25, 85)) +
    ggtitle(paste0("Permafrost T1m (degC) N=", nrow(pf_stats)))

ggplot(pf_stats, aes(x = PF_T1m_trend)) + 
  geom_histogram(bins = 30) +
  ggtitle("Permafrost trends (degC/decade)")

ggplot(pf_stats, 
       aes(x = PF_T1m_mean, y = Latitude, 
           color = PF_T1m_trend)) + 
  geom_point() +
  ggtitle("Trend versus latitude")

# Categorical (cooling/no sig. change/warming) trend map
pf_stats |> 
  mutate(Trend = if_else(PF_T1m_p.value >= 0.05, "None",
                         if_else(PF_T1m_trend < 0, 
                                 "Cooling", "Warming"))) ->
  results_categorical

ggplot(world) + 
    geom_sf() + 
    geom_point(data = results_categorical, 
               aes(x = Longitude, y = Latitude, color = Trend)) + 
  coord_sf(ylim = c(25, 85)) +
  scale_color_manual(values = c("blue", "darkgrey", "red")) +
  ggtitle("Permafrost temperature trends")
```

# Ancillary data

## WorldClim

```{r worldclim}
#| fig-width: 12

message("Doing WorldClim")
coords_worldclim <- coords
tavg <- worldclim_global(var = "tavg", path = "data-worldclim/", res = 5)
tavg_dat <- terra::extract(tavg, coords_worldclim[2:3])
coords_worldclim$WC_MAT <- rowMeans(tavg_dat[-1]) 

prec <- worldclim_global(var = "prec", 
                         path = "data-worldclim/", res = 5)
prec_dat <- terra::extract(prec, coords_worldclim[2:3])
coords_worldclim$WC_MAP <- rowSums(prec_dat[-1]) 

# Visualize

ggplot(coords_worldclim, aes(WC_MAT, WC_MAP)) + 
  geom_point(na.rm = TRUE, alpha = 0.5) + 
    ggtitle("WorldClim MAT and MAP")
```

## Soil and land cover

```{r soil-landcover}
#| fig-width: 12

message("Doing landcover...")
for(cov in c("trees", "grassland", "shrubs", "bare", "wetland", "moss")) {
  cover <- landcover(cov, path = "data-worldclim/")
  plot(cover, main = cov)
  cover_dat <- terra::extract(cover, coords_worldclim[2:3])
  coords_worldclim[[paste0("WC_", cov)]] <- cover_dat[,2]
}

message("Doing clay...")
clay <- soil_world("clay", depth = 5, stat = "mean", 
                   path = "data-worldclim/")
plot(clay, main = "clay")
clay_dat <- terra::extract(clay, coords_worldclim[2:3])
coords_worldclim$WC_clay <- clay_dat[,2]

```

## ERA5 climate

For each SRDB point we extract 2m air temperature, soil temperature
and water content (levels 1 and 2), precipitation, and low- and 
high-vegetation LAI from the 
[ERA5-Land monthly averaged data](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-land-monthly-means?tab=overview) 
(available from 1950 to present; I downloaded 1985-2024).

We extract 1995-2024, a 30-year period covering the 
permafrost thermal data range.

These data are *big* and operations are *slow*. 
We use three strategies to deal with this:

- Download individual-year `grib` files from the EMCWF Climate Data Store (a pain to download, but then extracting from them is relatively fast);

- Use a cache so we only process things once, hopefully;

- Parallelize.

```{r era5-settings}
# 1995-2024 is 30 year period that covers the permafrost thermal dataset
ERA5_YEARS <- 1995:2024

ERA5_SHORT_NAME <- read_csv("data-ERA5/data-ERA5/ERA5_short_names.csv", 
                            col_types = "cc")
ERA5_CACHE <- "caches/ERA5_cache.parquet"
```

Extract data for `r min(ERA5_YEARS)`-`r max(ERA5_YEARS)`.

```{r read-era5}
#| fig-height: 6
#| fig-width: 10

# Load cached data if available
if(file.exists(ERA5_CACHE)) {
  message("Reading cache file ", ERA5_CACHE)
  era5_cache <- arrow::read_parquet(ERA5_CACHE)
  message("\t", nrow(era5_cache), " rows")
} else {
  message("No ERA5 data cache found")
  era5_cache <- tibble(Year = numeric(0), 
                       Longitude = numeric(0),
                       Latitude = numeric(0))
}

# Helper function to read and process ERA5 data for one yr-lon-lat
read_era5 <- function(yr, lon, lat, ERA5_SHORT_NAME, my_coords) {
  # Construct year-specific file name; these were downloaded
  # from ECMWF and contain 2m temperature, soil temperature
  # levels 1 and 2,  volumetric soil water content levels 1 
  # and 2, total precipitation, and high and lower veg LAI
  f <- file.path("data-ERA5/data-ERA5", paste0("ERA5_", yr, ".grib"))
  if(!file.exists(f)) {
    message("\t", f, " does not exist; skipping")
    return(NULL)
  }
  x <- terra::rast(f)
  
  # Make sure the time period is what we're expecting
  if(!all(year(time(x)) == yr)) {
    stop("Year is supposed to be ", yr,
         " but file data dates are different!")
  }
  
  # Extract, reshape, clean up
  y <- terra::extract(x, my_coords[i,]) # this is the expensive step
  y |> 
    pivot_longer(-ID) |> 
    mutate(time = time(x), 
           Year = yr, 
           Longitude = lon, Latitude = lat,
           short_name = rep(ERA5_SHORT_NAME$short_name, 
                            length.out = n())) |> 
    select(Year, time, Longitude, Latitude, short_name, value) ->
    this_dat
  return(this_dat)
}

# Walk through the data and, if needed, 
# get data for each year (in parallel)
coords_unique <- coords |> distinct(Longitude, Latitude)
cache_hits <- 0
for(i in seq_len(nrow(coords_unique))) {
  
  lon <- coords_unique$Longitude[i]
  lat <- coords_unique$Latitude[i]
  # Isolate relevant cache data and compute years needed, if any
  era5_cache[era5_cache$Longitude == lon &
               era5_cache$Latitude == lat, "Year"] |> 
    pull() |> 
    unique() -> 
    unique_years
  years_needed <- setdiff(ERA5_YEARS, unique_years)
  
  if(length(years_needed) > 0) {
    message(i, "/", nrow(coords_unique), ": ",lon, " ", lat,
            " yn=", length(years_needed))
    
    # Extract, paralellizing over years, data for this lon/lat
    dat_list <- mclapply(years_needed,
                         FUN = read_era5, 
                         # yr is passed by default
                         lon = lon, lat = lat,
                         ERA5_SHORT_NAME = ERA5_SHORT_NAME,
                         my_coords = coords_unique,
                         mc.cores = 10)
    # ...and add to cache
    dat <- bind_rows(dat_list)
    era5_cache |> 
      bind_rows(dat) ->
      era5_cache
    message("\tAdding ", nrow(dat), " data rows to ", ERA5_CACHE)
    arrow::write_parquet(era5_cache, ERA5_CACHE)
  } else {
    cache_hits <- cache_hits + 1
  }
  
} # coords_unique loop

message("\tcache hits: ", cache_hits)
message("All done with ERA5 extractions")
```

Compute annual values, compute trends for Tair and Precip, 
and match individual-year values to SRDB year.

```{r era5-trends}
# At this point the cache data frame has all the needed data:
# 1995-2024 ERA5 for each unique SRDB data point.
# Reshape, summarise to annual (although this may change in
# the future), and join with SRDB data
era5_cache |> 
    mutate(time = as.Date(time)) |> 
    pivot_wider(names_from = "short_name") |> 
    group_by(Year, Longitude, Latitude) |>
    summarise(ERA5_Tair = round(mean(Tair) - 273.1, 2),
              ERA5_Tsoil_lev1 = round(mean(Tsoil_lev1) - 273.1, 2),
              #ERA5_Tsoil_lev2 = round(mean(Tsoil_lev2) - 273.1, 2),
              ERA5_VWC_lev1 = round(mean(VWC_lev1), 3),
              #ERA5_VWC_lev2 = round(mean(VWC_lev2), 3),
              ERA5_Precip = round(sum(Precip) * 1000, 2),
              ERA5_LAI_high = round(max(LAI_high), 2),
              ERA5_LAI_low = round(max(LAI_low), 2),
              .groups = "drop") ->
    era5_annual

# Compute trends - just Tair and Precip for now
era5_annual |> 
  select(Longitude, Latitude, Year, ERA5_Tair, ERA5_Precip) |> 
  pivot_longer(starts_with("ERA5")) |> 
  filter(!is.na(value)) |> 
  group_by(Longitude, Latitude, name) |> 
  reframe(tidy(lm(value * 10 ~ Year))) |> 
  filter(term == "Year") |> 
  select(Longitude, Latitude, name, trend = estimate, p.value) |> 
  pivot_wider(names_from = "name", values_from = c("trend", "p.value")) ->
  era5_trends

# A few sanity check plots
ggplot(world) + 
  geom_sf() + 
  geom_point(data = era5_trends, 
             aes(x = Longitude, y = Latitude, color = trend_ERA5_Tair),
             na.rm = TRUE) + 
  coord_sf(ylim = c(25, 85)) +
  ggtitle("ERA5 2m air temperature trends (degC/decade)")

ggplot(era5_trends, aes(x = trend_ERA5_Tair)) + 
  geom_histogram(bins = 30) +
  ggtitle("ERA5 Tair trends (degC/decade)")

ggplot(world) + 
  geom_sf() + 
  geom_point(data = era5_trends, 
             aes(x = Longitude, y = Latitude, color = trend_ERA5_Precip),
             na.rm = TRUE) + 
  coord_sf(ylim = c(25, 85)) +
  ggtitle("ERA5 precip trends (cm/decade)")
```


## ERA5 veg and soil

## Vegetation and soil type data

The vegetation and soil data are also from ERA5 but thankfully they're not time-varying, so fast and easy to extract.

```{r read-veg-soil-types}
#| warning: false
#| fig-width: 12

# Extract soil data for SRDB points
soil <- rast("data-ERA5/data-ERA5/ERA5_soil_type.grib")
plot(soil, main = "ERA5 soil type")
soil_dat <- terra::extract(soil, coords[2:3])
names(soil_dat) <- c("ID", "ERA5_soil_number")

# Attach names to numeric codes
soil_dat$ERA5_soil_number <- as.character(soil_dat$ERA5_soil_number) # for join below
soil_names <- tibble(strings = strsplit(units(soil), "; ")[[1]]) |>  
    separate(strings, into = c("ERA5_soil_number", "ERA5_soil_name"), 
             convert = TRUE, sep = "=")
soil_dat <- soil_dat |> left_join(soil_names, by = "ERA5_soil_number")

# Extract vegetation data
tvh <- rast("data-ERA5/data-ERA5/ERA5_tvh.grib")
plot(tvh, main = "ERA5 high vegetation type")
tvh_dat <- terra::extract(tvh, coords[2:3])
names(tvh_dat) <- c("ID", "ERA5_veg_high")
tvl <- rast("data-ERA5/data-ERA5/ERA5_tvl.grib")
plot(tvl, main = "ERA5 low vegetation type")
tvl_dat <- terra::extract(tvl, coords[2:3])
names(tvl_dat) <- c("ID", "ERA5_veg_low")

#lcc <- read_csv("data-ERA5/ECMWF_land_cover_codes.csv", col_types = "cc")
era5_soil_veg <- bind_cols(coords, soil_dat[-1], tvh_dat[-1], tvl_dat[-1])
```

## MODIS 8-day GPP

The [MODIS GPP MOD17A2HGF product](https://data.nasa.gov/dataset/modis-terra-gross-primary-productivity-gap-filled-8-day-l4-global-500m-sin-grid-v061-f3633)
is every eight days, 500 meter (m) resolution, and we sum it for an annual value.

S. Running, Zhao, M. 2021. MOD17A2HGF MODIS/Terra Gross Primary Productivity Gap-Filled 8-Day L4 Global 500 m SIN Grid V061. NASA EOSDIS Land Processes DAAC. https://doi.org/10.5067/MODIS/MOD17A2HGF.061

This process is *also* slow.

```{r modis-gpp}
#| fig-width: 12

# Load cached data if available
MODIS_GPP_CACHE <- "caches/MODIS_GPP_cache.parquet"
if(file.exists(MODIS_GPP_CACHE)) {
  message("Reading cache file ", MODIS_GPP_CACHE)
  cache <- arrow::read_parquet(MODIS_GPP_CACHE)
  message("\t", nrow(cache), " rows")
} else {
  message("No MODIS GPP data cache found")
  cache <- tibble(Year = numeric(0), 
                  calendar_date = character(0),
                  Longitude = numeric(0), Latitude = numeric(0),
                  MODIS_GPP = numeric(0))
}

# coords_yr (the unique combinations of location and time)
# was already calculated above
cache_hits <- 0
for(i in seq_len(nrow(coords_yr))) {
  yr <- coords_yr$Year[i]
  lon <- coords_yr$Longitude[i]
  lat <- coords_yr$Latitude[i]
  
  # Are data already in cache?
  cache |> 
    filter(Year == yr, Longitude == lon, Latitude == lat) ->
    dat
  if(nrow(dat) > 0) {
    cache_hits <- cache_hits + 1
  } else {
    # Use MODISTools to download the needed data
    if(cache_hits > 0) { 
      message("(", cache_hits, " cache hits)")
      cache_hits <- 0
    }
    
    message(i, "/", nrow(coords_yr), ": ", yr, " ", lon, " ", lat, appendLF = FALSE)
    message("\tNo data in cache; querying MODIS GPP products...")
    gpp <- NULL
    try({
      gpp <- mt_subset("MOD17A2HGF", 
                       band = "Gpp_500m", 
                       lat = lat, lon = lon, 
                       start = paste0(yr, "-01-01"), 
                       end = paste0(yr, "-12-31"))
    })
    if(is.null(gpp)) {
      gpp_dat <- tibble(Year = yr, 
                        calendar_date = NA_character_,
                        Longitude = lon, Latitude = lat, 
                        MODIS_GPP = NA_real_)
    } else {
      gpp_dat <- tibble(Year = yr, 
                        calendar_date = gpp$calendar_date,
                        Longitude = lon, Latitude = lat,
                        MODIS_GPP = gpp$value)
    }
    
    message("\tAdding ", nrow(gpp_dat), " data rows to cache ", MODIS_GPP_CACHE)
    cache <- bind_rows(cache, gpp_dat)
    arrow::write_parquet(cache, MODIS_GPP_CACHE)
  }
}

# Summarise cache, join, sanity check plot
cache |> 
  # 32767 is the 'fill value' for this data product -- no measurement
  # There are also some other values in the 32760s used for non-vegetated surfaces
  # see https://lpdaac.usgs.gov/documents/972/MOD17_User_Guide_V61.pdf
  mutate(MODIS_GPP = if_else(MODIS_GPP >= 32760, NA_real_, MODIS_GPP)) |> 
  group_by(Year, Longitude, Latitude) |> 
  summarise(MODIS_GPP = sum(MODIS_GPP) / 10.0, .groups = "drop") ->
  modis_gpp_annual

ggplot(world) + 
  geom_sf() + 
  geom_point(data = modis_gpp_annual, 
             aes(x = Longitude, y = Latitude, color = MODIS_GPP),
             na.rm = TRUE) + 
  ggtitle("MOD17A2HGF GPP (annual)")
```

# Combined QAQC

```{r combine-data}
message("Size before joins: ", nrow(srdb_sample), " x ", ncol(srdb_sample))

srdb_sample |> 
  left_join(pf_stats, 
            by = c("Record_number", "Longitude", "Latitude")) |> 
  left_join(coords_worldclim, 
            by = c("Record_number", "Longitude", "Latitude")) |> 
  left_join(era5_annual, 
            by = c("Longitude", "Latitude", "Year")) |> 
  left_join(era5_soil_veg, 
            by = c("Record_number", "Longitude", "Latitude")) |> 
  left_join(modis_gpp_annual, 
            by = c("Longitude", "Latitude", "Year")) |> 
  left_join(era5_trends,
            by = c("Longitude", "Latitude")) -> 
  srdb_joined

message("Size after joins: ", nrow(srdb_joined), " x ", ncol(srdb_joined))

```

NA PF_GST_mean records: `r sum(is.na(srdb_joined$PF_GST_mean))`
NA WC_MAT records: `r sum(is.na(srdb_joined$WC_MAT))`
NA ERA_Tair records: `r sum(is.na(srdb_joined$ERA5_Tair))`

```{r qaqc}
srdb_joined |> 
    filter(Rs_annual > 0) |> 
    transmute(`sqrt(Rs_growingseason)` = sqrt(Rs_growingseason),
              `sqrt(Rs_annual)` = sqrt(Rs_annual), 
              `sqrt(Rh_annual)` = sqrt(Rh_annual)) |> 
    pairs()

# GPP
srdb_joined |> 
  filter(!is.na(MODIS_GPP), !is.na(SRDB_GPP)) |> 
  ggplot(aes(SRDB_GPP, MODIS_GPP)) +
  geom_point() +
  geom_abline()

# Climate
srdb_joined |> 
  filter(!is.na(ERA5_Tair), !is.na(WC_MAT)) |> 
  ggplot(aes(WC_MAT, ERA5_Tair)) +
  geom_point() +
  geom_abline()

srdb_joined |> 
  filter(!is.na(ERA5_Precip), !is.na(WC_MAP)) |> 
  ggplot(aes(WC_MAP, ERA5_Precip * 10)) +
  geom_point() +
  geom_abline()
```

Write data to file for next step in the analysis.

```{r write-data}
OUTFILE <- "srdb_joined_data.parquet"
message("Writing ", OUTFILE)
write_parquet(srdb_joined, OUTFILE)
```

# Reproducibility

```{r reproducibility}
#| echo: false

sessionInfo()
```
