---
title: "1-permafrost-data-prep"
author: "bbl"
format:
  html:
    embed-resources: true
    code-fold: true
toc: true
editor: visual
---

**TODO:**

- Add MODIS GPP (from gsbe)
- Add some QA/QC and summary info to bottom of this
- There's a lot of NA MAT and MAP from WorldClim (see next bullet)
- Do I want to change to ERA5? Probably b/c that will let me get year-specific info
- Try global Random Forest model with permafrost flag
- CLEAN UP GRACE'S STUFF AND MERGE AND UPDATE
- Deal with the remaining papers in my to-process folder, prioritizing high latitude ones

This file does prep work for the main analysis asking whether permafrost *presence* and *thermal state* affect the measured soil-to-atmosphere CO2 flux ("soil respiration").

```{r setup}
#| include: false

# Data processing and visualization packages
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())
library(lubridate)

# File IO packages
library(ncdf4)
library(arrow)
library(readr)

# Statistics and spatial packages
library(broom)
library(sf)
library(terra)

# Data packages
library(geodata)
library(rnaturalearth)
```

## SRDB

The Soil Respiration Database (SRBD) is described by [Jian et al. (2021)](https://essd.copernicus.org/articles/13/255/2021/), but the version we use here has had extensive updates since then.

```{r read-srdb}
#| fig-width: 12

read_csv("data-srdb/srdb-data.csv", show_col_types = FALSE) %>% 
    select(Record_number, Study_number, Author, 
           Year = Study_midyear, YearsOfData, 
           Latitude, Longitude, Elevation, Manipulation, 
           Biome, Ecosystem_state, 
           Rs_annual, Rh_annual, Rs_growingseason,
           Ecosystem_type,
           # intercomparison columns -- will not be used afterwards
           SRDB_study_temp = Study_temp, SRDB_study_precip = Study_precip, 
           SRDB_GPP = GPP, SRDB_ANPP = ANPP, SRDB_leaf_habit = Leaf_habit) %>% 
    filter(Manipulation == "None") %>% 
    filter(!Ecosystem_type %in% c("Agriculture", "Bare", "Urban")) %>% 
    filter(!is.na(Longitude), !is.na(Latitude), !is.na(Year)) %>% 
    filter(!is.na(Rs_growingseason) | !is.na(Rs_annual) | !is.na(Rh_annual)) %>% 
    select(-Elevation, -Manipulation) %>% 
    mutate(Longitude = round(Longitude, 4),
           Latitude = round(Latitude, 4),
           Year = floor(Year)) ->
    srdb

message(nrow(srdb), " rows of data")

world <- ne_coastline()
srdb_annual <- srdb %>% filter(is.na(Rs_annual))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_annual, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rs_annual N=", nrow(srdb_annual)))

srdb_rh_annual <- srdb %>% filter(!is.na(Rh_annual))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_rh_annual, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rh_annual N=", nrow(srdb_rh_annual)))

srdb_growingseason <- srdb %>% filter(is.na(Rs_growingseason))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_growingseason, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rs_growingseason N=", nrow(srdb_growingseason)))
```

```{r srdb-prep}
#| fig-width: 12

# Subsample SRDB data for quick testing
set.seed(12)
srdb %>%
  filter(Latitude > 50) |> 
    slice_sample(n = 100) ->
    srdb_sample
srdb_sample <- srdb

srdb_sample %>% 
    select(Record_number, Longitude, Latitude) ->
    coords

srdb_sample %>% 
  distinct(Year, Longitude, Latitude) ->
  coords_yr
```

## ESA-CCI permafrost

Our data source for permafrost temperatures is Westermann et al.: ESA permafrost Climate Change Initiative (permafrost_cci): Permafrost ground temperature for the Northern Hemisphere, v5.0, <https://doi.org/10.5285/5675B0BE944F45A8AF0E7DDBEB47A011>, 2025.

```{r read-essacci-permafrost}
#| fig-height: 6
#| fig-width: 10

PERMAFROST_CACHE <- "caches/permafrost_cache.parquet"
if(file.exists(PERMAFROST_CACHE)) {
  message("Reading data from permafrost cache...")
  pf_cache <- arrow::read_parquet(PERMAFROST_CACHE) 
  message("\t(", nrow(pf_cache), " rows)")
} else {
  message("No permafrost cache file found")
  pf_cache <- tibble(Record_number = numeric(0), 
                     Longitude = numeric(0), 
                     Latitude = numeric(0))
}

# Isolate SRDB lon/lats NOT in the cache
coords |> 
  anti_join(pf_cache, by = c("Record_number", "Longitude", "Latitude")) ->
  coords_no_data
if(nrow(coords_no_data) > 0) {
  message("There are ", nrow(coords_no_data), 
          " coordinates that need data")
  
  results_list <- list()

  files <- list.files("data-permafrost_cci//", 
                      full.names = TRUE,
                      pattern = "\\.nc$",
                      recursive = TRUE)
 
  for(fn in files) {
    # Trying a simple terra::rast() with these data gives
    # Error: [rast] could not load multidimensional data. 
    # So use ncdf4 to load; transpose and flip; and convert
    # to a raster, specifying the spatial extent and CRS
    message("Opening ", basename(fn))
    nc <- nc_open(fn)
    
    ncvars_to_read <- c("GST", "GST_uncertainty", "T1m", "T1m_uncertainty")
    for(v in ncvars_to_read) {
      message("\tGetting ", v)
      dat_matrix <- ncvar_get(nc, varid = v)
      
      # Transpose and flip the data
      dat_matrix <- t(dat_matrix)
      dat_matrix <- dat_matrix[dim(dat_matrix)[1]:1,]
      # Convert to spatial raster, specifying lon/lat extent
      dat_rast <- rast(dat_matrix, 
                       extent = c(-180, 180, 25, 85), 
                       crs="+proj=longlat +datum=WGS84")
      message("\tExtracting points...")
      xdat <- terra::extract(dat_rast,
                             coords_no_data[c("Longitude", "Latitude")])
      coords_no_data[paste0("var_", v)] <- xdat[2]
    }
    rm(dat_matrix, dat_rast) # clean up; these are big
    
    # The year is in the filename; extract it (a bit hackily)
    yr <- strsplit(basename(fn), "-", fixed = TRUE)[[1]][7]
    coords_no_data$Year <- as.numeric(yr)
    results_list[[yr]] <- coords_no_data
    
    nc_close(nc)
  }
  
  bind_rows(results_list) |> 
    bind_rows(pf_cache) ->
    pf_cache
  
  message("Writing cache file (", nrow(pf_cache), " rows)...")
  arrow::write_parquet(pf_cache, PERMAFROST_CACHE)
}


p <- ggplot(filter(pf_cache, !is.na(var_T1m)), 
            aes(Year, var_T1m, color = Latitude, group = Record_number)) +
  geom_line(na.rm = TRUE) 
print(p)

pf_cache |> 
  filter(!is.na(var_T1m)) |> 
  distinct(Record_number, Year, Latitude, Longitude) ->
  pf_points
```

OK!

```{r permafrost-stats}
#| fig-width: 12

message("Computing means and trends...")
pf_cache |> 
  na.omit() |> 
  select(Record_number, Longitude, Latitude, Year, starts_with("var_")) |> 
  select(-ends_with("uncertainty")) ->
  results_non_uncertainty

# Compute means
results_non_uncertainty |> 
  pivot_longer(starts_with("var_")) |> 
  group_by(Record_number, Longitude, Latitude, name) |> 
  summarise(value = mean(value), .groups = "drop") |> 
  mutate(name = paste0(name, "_mean")) |> 
  pivot_wider() ->
  results_mean

# Compute trends
results_non_uncertainty |> 
  pivot_longer(starts_with("var_")) |> 
  filter(!is.na(value)) |> 
  group_by(Record_number, Longitude, Latitude, name) |> 
  reframe(lm(value ~ Year) |> tidy()) |> 
  filter(term == "Year") |> 
  mutate(estimate = estimate * 10) |> # change to degrees per decade
  select(Record_number, Longitude, Latitude, name, estimate, p.value) ->
  trends

trends |> 
  select(Record_number, Longitude, Latitude, name, estimate) |> 
  mutate(name = paste0(name, "_trend")) |> 
  pivot_wider(values_from = "estimate") ->
  trends_est
trends |> 
  select(Record_number, Longitude, Latitude, name, p.value) |> 
  mutate(name = paste0(name, "_p.value")) |> 
  pivot_wider(values_from = "p.value") ->
  trends_p

results_mean |> 
  left_join(trends_est, 
            by = c("Record_number", "Longitude", "Latitude")) |> 
  left_join(trends_p, 
            by = c("Record_number", "Longitude", "Latitude")) ->
  pf_stats

# Visualize

ggplot(world) + 
    geom_sf() + 
    geom_point(data = pf_stats, 
               aes(x = Longitude, y = Latitude, color = var_T1m_mean)) + 
  coord_sf(ylim = c(25, 85)) +
    ggtitle(paste0("Permafrost T1m (degC) N=", nrow(pf_stats)))

ggplot(world) + 
    geom_sf() + 
    geom_point(data = pf_stats, 
               aes(x = Longitude, y = Latitude, color = var_T1m_trend)) + 
  coord_sf(ylim = c(25, 85)) +
    ggtitle(paste0("Permafrost var_T1m_trend (degC/decade) N=", nrow(pf_stats)))

```

```{r trend-visualization}
#| fig-width: 12

ggplot(pf_stats, 
       aes(x = var_T1m_mean, y = Latitude, 
           color = var_T1m_trend)) + 
  geom_point() +
  ggtitle("Trend versus latitude")

# Categorical (cooling/no sig. change/warming) trend map
pf_stats |> 
  mutate(Trend = if_else(var_T1m_p.value >= 0.05, "None",
                         if_else(var_T1m_trend < 0, 
                                 "Cooling", "Warming"))) ->
  results_categorical
ggplot(world) + 
    geom_sf() + 
    geom_point(data = results_categorical, 
               aes(x = Longitude, y = Latitude, color = Trend)) + 
  coord_sf(ylim = c(25, 85)) +
  scale_color_manual(values = c("blue", "darkgrey", "red")) +
  ggtitle("Permafrost temperature trends")
```

## WorldClim

```{r worldclim}
#| fig-width: 12

message("Doing WorldClim")
coords_worldclim <- coords
tavg <- worldclim_global(var = "tavg", path = "data-worldclim/", res = 5)
tavg_dat <- terra::extract(tavg, coords_worldclim[2:3])
coords_worldclim$WC_MAT <- rowMeans(tavg_dat[-1]) 

prec <- worldclim_global(var = "prec", 
                         path = "data-worldclim/", res = 5)
prec_dat <- terra::extract(prec, coords_worldclim[2:3])
coords_worldclim$WC_MAP <- rowSums(prec_dat[-1]) 

# Visualize

ggplot(coords_worldclim, aes(WC_MAT, WC_MAP)) + 
  geom_point(na.rm = TRUE, alpha = 0.5) + 
    ggtitle("WorldClim MAT and MAP")
```

## Soil and land cover

```{r soil-landcover}
#| fig-width: 12

message("Doing landcover...")
for(cov in c("trees", "grassland", "shrubs", "bare", "wetland", "moss")) {
  cover <- landcover(cov, path = "data-worldclim/")
  plot(cover, main = cov)
  cover_dat <- terra::extract(cover, coords_worldclim[2:3])
  coords_worldclim[[paste0("cover_", cov)]] <- cover_dat[,2]
}

message("Doing clay...")
clay <- soil_world("clay", depth = 5, stat = "mean", 
                   path = "data-worldclim/")
clay_dat <- terra::extract(clay, coords_worldclim[2:3])
coords_worldclim$Clay <- clay_dat[,2]

# Visualize

ggplot(world) + 
    geom_sf() + 
    geom_point(data = coords_worldclim, 
               aes(x = Longitude, y = Latitude, color = Clay)) + 
    ggtitle("WorldClim Clay")
```

## ERA5

For each SRDB point we extract 2m air temperature, soil temperature and water content (levels 1 and 2), precipitation, and high-vegetation LAI from the [ERA5-Land monthly averaged data](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-land-monthly-means?tab=overview) (available from 1950 to present; I downloaded 1985-2024).

These data are *big* and operations are *slow*. We use two strategies to deal with this:

- Download individual-year `grib` files from the EMCWF Climate Data Store (a pain to download, but then extracting from them is relatively fast); and

- Use a cache so we only process things once, hopefully

```{r read-era5}
#| fig-height: 6
#| fig-width: 10

# Load cached data if available
ERA5_SHORT_NAME <- read_csv("data-ERA5/data-ERA5/ERA5_short_names.csv", col_types = "cc")
ERA5_CACHE <- "caches/ERA5_data_cache.parquet"
if(file.exists(ERA5_CACHE)) {
    message("Reading cache file ", ERA5_CACHE)
    cache <- arrow::read_parquet(ERA5_CACHE)    
} else {
    message("No ERA5 data cache found")
    cache <- tibble(Year = numeric(0), Longitude = numeric(0), Latitude = numeric(0))
}

# Get distinct location combinations
# We only get ERA5 data for locations with permafrost thermal data
pf_cache |> 
  filter(!is.na(var_GST)) |> 
    distinct(Longitude, Latitude) ->
    pf_lonlat

cache_hits <- 0
for(i in seq_len(nrow(pf_lonlat))) {
  lon <- pf_lonlat$Longitude[i]
  lat <- pf_lonlat$Latitude[i]
  
  for(yr in 1995:2024) {
    # Are data already in cache?
    cache %>% 
      filter(Year == yr, Longitude == lon, Latitude == lat) ->
      dat
    if(nrow(dat) > 0) {
      cache_hits <- cache_hits + 1
      #message("\t", nrow(dat), " data rows are in cache")
    } else {
      if(cache_hits > 0) { 
        message("(", cache_hits, " cache hits)")
        cache_hits <- 0
      }
      message(i, "/", nrow(pf_lonlat), ": ", yr, " ", lon, " ", lat, appendLF = FALSE)
      # Construct year-specific file name; these were downloaded from ECMWF
      # and contain 2m temperature, soil temperature levels 1 and 2, 
      # volumetric soil water content levels 1 and 2, total precipitation,
      # and LAI of high and low vegetation
      f <- file.path("data-ERA5/data-ERA5", paste0("ERA5_", yr, ".grib"))
      if(!file.exists(f)) {
        message("\t", f, " does not exist; skipping")
        next
      }
      message("\tNo data in cache; extracting from ", f, "...")
      x <- terra::rast(f)
      
      # Make sure the time period is what we're expecting
      if(!all(year(time(x)) == yr)) {
        stop("Year is supposed to be ", yr, " but file data dates are different!")
      }
      
      # Extract, reshape, clean up
      y <- terra::extract(x, pf_lonlat[i,]) # this is the expensive step
      y %>% 
        pivot_longer(-ID) %>% 
        mutate(time = time(x), 
               Year = yr, 
               Longitude = lon, Latitude = lat,
               short_name = rep(ERA5_SHORT_NAME$short_name, length.out = n())) %>% 
        select(Year, time, Longitude, Latitude, short_name, value) ->
        dat
      cache <- bind_rows(cache, dat)
      message("\tAdding ", nrow(dat), " data rows to cache ", ERA5_CACHE, "...")
      arrow::write_parquet(cache, ERA5_CACHE)
    }
  } # yr loop
}

# At this point the cache data frame has all the needed data:
# 1995-2024 ERA5 for each point permafrost-region point.
# Reshape, summarise to annual (although this may change in
# the future), and join with SRDB data
message("Summarizing and joining with SRDB data...")
cache %>% 
    mutate(time = as.Date(time)) %>% 
    pivot_wider(names_from = "short_name") %>% 
    group_by(Year, Longitude, Latitude) %>%
    summarise(ERA5_Tair = round(mean(Tair) - 273.1, 2),
              ERA5_Tsoil_lev1 = round(mean(Tsoil_lev1) - 273.1, 2),
              #ERA5_Tsoil_lev2 = round(mean(Tsoil_lev2) - 273.1, 2),
              ERA5_VWC_lev1 = round(mean(VWC_lev1), 3),
              #ERA5_VWC_lev2 = round(mean(VWC_lev2), 3),
              ERA5_Precip = round(sum(Precip) * 1000, 2),
              ERA5_LAI_high = round(max(LAI_high), 2),
              ERA5_LAI_low = round(max(LAI_low), 2),
              .groups = "drop") ->
    cache_summary

# Compute trends - just Tair and Precip for now
cache_summary |> 
  select(Longitude, Latitude, Year, ERA5_Tair, ERA5_Precip) |> 
  pivot_longer(starts_with("ERA5")) |> 
  filter(!is.na(value)) |> 
  group_by(Longitude, Latitude, name) |> 
  reframe(tidy(lm(value * 10 ~ Year))) |> 
  filter(term == "Year") |> 
  select(Longitude, Latitude, name, trend = estimate, p.value) |> 
  pivot_wider(names_from = "name", values_from = c("trend", "p.value")) ->
  cache_trends

srdb_sample %>% 
    left_join(cache_summary, by = c("Year", "Longitude", "Latitude")) ->
    srdb_sample

# A few sanity check plots
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_sample, 
               aes(x = Longitude, y = Latitude, color = ERA5_Tair),
               na.rm = TRUE) + 
    ggtitle("SRDB with ERA5 2m air temperature")

ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_sample, 
               aes(x = Longitude, y = Latitude, color = ERA5_Precip),
               na.rm = TRUE) + 
    ggtitle("SRDB with ERA5 total precipitation")

ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_sample, 
               aes(x = Longitude, y = Latitude, color = ERA5_LAI_high),
               na.rm = TRUE) + 
    ggtitle("SRDB with ERA5 max high LAI")
```

## Combined QAQC

```{r combine-data}
message("Size before joins: ", nrow(srdb_sample))

srdb_sample |> 
  left_join(pf_stats, by = join_by("Record_number", "Latitude", "Longitude")) |> 
  left_join(coords_worldclim, by = join_by("Record_number", "Latitude", "Longitude")) -> 
  srdb_joined

message("Size after joins: ", nrow(srdb_joined))
```

Writing data to file for next step in the analysis.

```{r write-data}
OUTFILE <- "srdb_joined_data.parquet"
message("Writing ", OUTFILE)
write_parquet(srdb_joined, OUTFILE)
```

## Reproducibility

```{r}
#| echo: false

sessionInfo()
```
