---
title: "1-permafrost-data-prep"
author: "bbl"
format:
  html:
    embed-resources: true
    code-fold: true
toc: true
editor: visual
---

```{r setup}
#| include: false

# Data processing and visualization packages
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())
library(lubridate)
library(DT)

# File IO packages
library(ncdf4)
library(arrow)
library(readr)

# Statistics and spatial packages
library(broom)
library(sf)
library(terra)

# Data packages
library(geodata)
library(rnaturalearth)
```

## SRDB

The Soil Respiration Database (SRBD) is described by [Jian et al. (2021)](https://essd.copernicus.org/articles/13/255/2021/), but the version we use here has had extensive updates since then.

```{r read-srdb}
#| fig-width: 12

read_csv("data-srdb/srdb-data.csv", show_col_types = FALSE) %>% 
    select(Record_number, Study_number, Author, 
           Study_midyear, YearsOfData, 
           Latitude, Longitude, Elevation, Manipulation, 
           Biome, Ecosystem_state, 
           Rs_annual, Rh_annual, Rs_growingseason,
           Ecosystem_type,
           # intercomparison columns -- will not be used afterwards
           SRDB_study_temp = Study_temp, SRDB_study_precip = Study_precip, 
           SRDB_GPP = GPP, SRDB_ANPP = ANPP, SRDB_leaf_habit = Leaf_habit) %>% 
    filter(Manipulation == "None") %>% 
    filter(!Ecosystem_type %in% c("Agriculture", "Bare", "Urban")) %>% 
    filter(!is.na(Longitude), !is.na(Latitude), !is.na(Study_midyear)) %>% 
    filter(!is.na(Rs_growingseason) | !is.na(Rs_annual) | !is.na(Rh_annual)) %>% 
    select(-Elevation, -Manipulation) %>% 
    mutate(Longitude = round(Longitude, 4),
           Latitude = round(Latitude, 4)) ->
    srdb

message(nrow(srdb), " rows of data")

world <- ne_coastline()
srdb_annual <- srdb %>% filter(is.na(Rs_annual))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_annual, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rs_annual N=", nrow(srdb_annual)))

srdb_rh_annual <- srdb %>% filter(!is.na(Rh_annual))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_rh_annual, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rh_annual N=", nrow(srdb_rh_annual)))

srdb_growingseason <- srdb %>% filter(is.na(Rs_growingseason))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_growingseason, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rs_growingseason N=", nrow(srdb_growingseason)))
```

```{r srdb-prep}
#| fig-width: 12

# Subsample SRDB data for quick testing
set.seed(12)
srdb %>%
  filter(Latitude > 50) |> 
    slice_sample(n = 100) ->
    srdb_sample
srdb_sample <- srdb

srdb_sample %>% 
    select(Record_number, Longitude, Latitude) ->
    coords

srdb_sample %>% 
  distinct(Study_midyear, Longitude, Latitude) ->
  coords_yr
```

## ESA-CCI permafrost

Our data source for permafrost temperatures is Westermann et al.: ESA permafrost Climate Change Initiative (permafrost_cci): Permafrost ground temperature for the Northern Hemisphere, v5.0, https://doi.org/10.5285/5675B0BE944F45A8AF0E7DDBEB47A011, 2025.

```{r read-essacci-permafrost}
#| fig-height: 6
#| fig-width: 10

PERMAFROST_CACHE <- "caches/permafrost_cache.parquet"
if(file.exists(PERMAFROST_CACHE)) {
  message("Reading data from permafrost cache...")
  pf_cache <- arrow::read_parquet(PERMAFROST_CACHE) 
  message("\t(", nrow(pf_cache), " rows)")
} else {
  message("No permafrost cache file found")
  pf_cache <- tibble(Record_number = numeric(0), 
               Longitude = numeric(0), Latitude = numeric(0))
}

# Isolate SRDB lon/lats NOT in the cache
coords |> 
  anti_join(pf_cache, by = c("Record_number", "Longitude", "Latitude")) ->
  coords_no_data
if(nrow(coords_no_data) > 0) {
  message("There are ", nrow(coords_no_data), " coordinates that need data")
  
  results_list <- list()

  files <- list.files("data-permafrost_cci//", 
                      full.names = TRUE,
                      pattern = "\\.nc$",
                      recursive = TRUE)
 
  for(fn in files) {
    # Trying a simple terra::rast() with these data gives
    # Error: [rast] could not load multidimensional data. 
    # So use ncdf4 to load; transpose and flip; and convert
    # to a raster, specifying the spatial extent and CRS
    message("Opening ", basename(fn))
    nc <- nc_open(fn)
    
    ncvars_to_read <- c("T1m")
    #ncvars_to_read <- c("GST", "GST_uncertainty", "T1m", "T1m_uncertainty")
    for(v in ncvars_to_read) {
      message("\tGetting ", v)
      dat_matrix <- ncvar_get(nc, varid = v)
      
      # Transpose and flip the data
      dat_matrix <- t(dat_matrix)
      dat_matrix <- dat_matrix[dim(dat_matrix)[1]:1,]
      # Convert to spatial raster, specifying lon/lat extent
      dat_rast <- rast(dat_matrix, 
                       extent = c(-180, 180, 25, 85), 
                       crs="+proj=longlat +datum=WGS84")
      message("\tExtracting points...")
      xdat <- terra::extract(dat_rast,
                             coords_no_data[c("Longitude", "Latitude")])
      coords_no_data[paste0("var_", v)] <- xdat[2]
    }
    rm(dat_matrix, dat_rast) # clean up; these are big
    
    # The year is in the filename; extract it (a bit hackily)
    yr <- strsplit(basename(fn), "-", fixed = TRUE)[[1]][7]
    coords_no_data$Year <- as.numeric(yr)
    results_list[[yr]] <- coords_no_data
    
    nc_close(nc)
  }
  
  bind_rows(results_list) |> 
    bind_rows(pf_cache) ->
    pf_cache
  
  message("Writing cache file (", nrow(results), " rows)...")
  arrow::write_parquet(pf_cache, PERMAFROST_CACHE)
}


p <- ggplot(filter(pf_cache, !is.na(var_T1m)), 
            aes(Year, var_T1m, color = Latitude, group = Record_number)) +
  geom_line(na.rm = TRUE) 
print(p)

pf_cache |> 
  filter(!is.na(var_T1m)) |> 
  distinct(Record_number, Year, Latitude, Longitude) ->
  pf_points
```

OK!

```{r permafrost-stats}
#| fig-width: 12

message("Computing means and trends...")
pf_cache |> 
  na.omit() |> 
  select(Record_number, Longitude, Latitude, Year, starts_with("var_")) |> 
  select(-ends_with("uncertainty")) ->
  results_non_uncertainty

# Compute means
results_non_uncertainty |> 
  pivot_longer(starts_with("var_")) |> 
  group_by(Record_number, Longitude, Latitude, name) |> 
  summarise(value = mean(value), .groups = "drop") |> 
  mutate(name = paste0(name, "_mean")) |> 
  pivot_wider() ->
  results_mean

# Compute trends
results_non_uncertainty |> 
  pivot_longer(starts_with("var_")) |> 
  filter(!is.na(value)) |> 
  group_by(Record_number, Longitude, Latitude, name) |> 
  reframe(lm(value ~ Year) %>% tidy()) |> 
  filter(term == "Year") |> 
  mutate(estimate = estimate * 10) |> # degrees per decade
  select(Record_number, Longitude, Latitude, name, estimate, p.value) ->
  trends

trends |> 
  select(Record_number, Longitude, Latitude, name, estimate) |> 
  mutate(name = paste0(name, "_trend")) |> 
  pivot_wider(values_from = "estimate") ->
  trends_est
trends |> 
  select(Record_number, Longitude, Latitude, name, p.value) |> 
  mutate(name = paste0(name, "_p.value")) |> 
  pivot_wider(values_from = "p.value") ->
  trends_p

results_mean |> 
  left_join(trends_est, by = c("Record_number", "Longitude", "Latitude")) |> 
  left_join(trends_p, c("Record_number", "Longitude", "Latitude")) ->
  pf_stats

ggplot(world) + 
    geom_sf() + 
    geom_point(data = pf_stats, 
               aes(x = Longitude, y = Latitude, color = var_T1m_mean)) + 
  coord_sf(ylim = c(25, 85)) +
    ggtitle(paste0("Permafrost T1m (degC) N=", nrow(pf_stats)))

ggplot(world) + 
    geom_sf() + 
    geom_point(data = pf_stats, 
               aes(x = Longitude, y = Latitude, color = var_T1m_trend)) + 
  coord_sf(ylim = c(25, 85)) +
    ggtitle(paste0("Permafrost var_T1m_trend (degC/decade) N=", nrow(pf_stats)))

```

```{r trend-visualization}
#| fig-width: 12

ggplot(pf_stats, 
       aes(x = var_T1m_mean, y = Latitude, 
           color = var_T1m_trend)) + 
  geom_point() +
  ggtitle("Trend versus latitude")

# Categorical (cooling/none/warming) trend map
pf_stats |> 
  mutate(Trend = if_else(var_T1m_p.value >= 0.05, "None",
                         if_else(var_T1m_trend < 0, 
                                 "Cooling", "Warming"))) ->
  results_categorical
ggplot(world) + 
    geom_sf() + 
    geom_point(data = results_categorical, 
               aes(x = Longitude, y = Latitude, color = Trend)) + 
  coord_sf(ylim = c(25, 85)) +
  scale_color_manual(values = c("blue", "darkgrey", "red")) +
  ggtitle("Permafrost temperature trends")
```

## WorldClim and allies

```{r geodata}
message("Doing WorldClim")
coords_worldclim <- coords
tavg <- worldclim_global(var = "tavg", path = "data-worldclim/", res = 5)
tavg_dat <- terra::extract(tavg, coords_worldclim[2:3])
coords_worldclim$MAT_wc <- rowMeans(tavg_dat[-1]) 

prec <- worldclim_global(var = "prec", 
                         path = "data-worldclim/", res = 5)
prec_dat <- terra::extract(prec, coords_worldclim[2:3])
coords_worldclim$MAP_wc <- rowSums(prec_dat[-1]) 

message("Doing landcover...")
for(cov in c("trees", "grassland", "shrubs", "bare", "wetland", "moss")) {
  cover <- landcover(cov, path = "data-worldclim/")
  plot(cover, main = cov)
  cover_dat <- terra::extract(cover, coords_worldclim[2:3])
  coords_worldclim[[paste0("cover_", cov)]] <- cover_dat[,2]
}

message("Doing clay...")
clay <- soil_world("clay", depth = 5, stat = "mean", 
                   path = "data-worldclim/")
clay_dat <- terra::extract(clay, coords_worldclim[2:3])
coords_worldclim$Clay <- clay_dat[,2]

# Visualize

ggplot(world) + 
    geom_sf() + 
    geom_point(data = coords_worldclim, 
               aes(x = Longitude, y = Latitude, color = Clay)) + 
    ggtitle("WorldClim Clay")

ggplot(coords_worldclim, aes(MAT_wc, MAP_wc)) + 
  geom_point(na.rm = TRUE) + 
    ggtitle("WorldClim MAT and MAP")

```

```{r combine-data}

srdb_sample |> 
  left_join(pf_stats, by = join_by("Record_number", "Latitude", "Longitude")) |> 
  left_join(coords_worldclim, by = join_by("Record_number", "Latitude", "Longitude")) -> 
  srdb_joined

```


```{r write-data}
OUTFILE <- "srdb_joined_data.parquet"
message("Writing ", OUTFILE)
write_parquet(srdb_joined, OUTFILE)

DT::datatable(srdb_joined)
```

## Reproducibility

```{r}
#| echo: false

sessionInfo()
```
